# Join is an overlap

Objective: learn that a "join" is like an overlap.

We start with a quick example explaining why we use functions called
`join`.

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(randomNames)
```

Let's set up a simulation where we have 17 classmates (not counting
ourselves) in a school of 118. Suppose every member of the class has
12 friends outside of class. 

We want to find out, for each classmate, how many friends we have in
common and also, of our shared friends, what is a typical birthday.
Let's define typical as the median birthday of our shared friends.

```{r}
m <- 17 # size of class
n <- 100 # size of others in school
nf <- 12 # number of friends outside class
set.seed(5)
people <- randomNames(m + n,
                      name.order="first.last",
                      name.sep=" ")
```

We define the population of potential friends (those outside the
class) as `pop`, and our 12 friends are in `friends`. Finally, we
define the people that are in our class as `class_people`.

```{r}
pop <- data.frame(person=people[-(1:m)],
                  birthday=sample(31, n, TRUE))
friends <- data.frame(friend=sample(people, nf))
class_people <- people[1:m]
```

The following sets up a data.frame, where each row gives, for a given
classmate, their friends, and the friends' birthdays.

```{r}
loop <- rep(1:m, each=nf)
idx <- as.vector(replicate(m, sample(n, nf)))
class <- data.frame(classmate=class_people[loop],
                    friend=pop$person[idx],
                    birthday=pop$birthday[idx])
```

Sort this by classmate alphabetically, and make classmate into a
factor. This last step is important, as it will help us to keep track
of the classmates for whom we share _no_ friends outside of class.

```{r}
class <- class %>%
  arrange(classmate) %>%
  mutate(classmate=factor(classmate))
```

We perform an `inner_join` by `"friend"`. This just means we look for
classmates where we have overlapping friends, and we drop the rows
where we don't share any friends. "Inner" refers to the fact that we
are keeping the overlap in the _inside_ of two intersecting circles.
Note that the join operation brings along the metadata (extra data)
about the friends' birthdays.

```{r}
shared_friends <- friends %>%
  inner_join(class, by="friend") %>%
  arrange(classmate, friend) %>%
  relocate(classmate) # classmate to 1st column
shared_friends
```

Lastly, we perform some summarization: compute the number of friends
in common with `n()` and the median birthday of shared friends. The
`complete` call at the end fills in 0 for those classmates for
whom we share no friends (here, the use of `factor` earlier becomes
relevant).

```{r}
shared_friends %>%
  group_by(classmate) %>%
  summarize(in_common=n(), median_bday=median(birthday)) %>%
  complete(classmate, fill=list(count=0))
```
