# Tidy gene plots

Objective: for a region of the genome, find peaks near the TSS of genes
and then plot their signal strength per gene, stratifying by the tissue
origin of the peak.

We start by loading the pre-downloaded peaks ranges:

```{r}
load("data/peaks.rda")
```

Likewise, we want to use hg19 genes again to match the hg19 peaks:

```{r message=FALSE}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(org.Hs.eg.db)
g <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
```

Adding gene symbols:

```{r message=FALSE}
library(plyranges)
g <- g %>%
  mutate(gene_name = mapIds(org.Hs.eg.db, 
                            gene_id, "SYMBOL", "ENTREZID"))
```

Find a region of the genome near a kidney-specific gene:

```{r}
g %>% 
  filter(gene_name == "UMOD")
region <- data.frame(
  seqnames="chr16", 
  start=20e6,
  end=21e6) %>%
  as_granges()
```

Combine the bladder and kidney peaks, and select certain columns:

```{r}
pks <- bind_ranges(bladder=bladder_pks,
                   kidney=kidney_pks,
                   .id="tissue") %>%
  select(signal=signalValue, tissue)
```

Finally, we perform the overlap join, locating peaks within 10kb of 
the TSS of the gene.

```{r}
g_with_pks <- g %>%
  anchor_5p() %>%
  mutate(width=1) %>%
  filter_by_overlaps(region) %>%
  join_overlap_inner(pks, maxgap=1e4)
g_with_pks$tissue %>% table()
```

We can construct a faceted set of boxplots:

```{r peaks-near-genes, message=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)
g_with_pks %>%
  select(gene_name, signal, tissue, .drop_ranges=TRUE) %>%
  as_tibble() %>%
  ggplot(aes(tissue, signal)) +
  geom_boxplot() + 
  facet_wrap(~gene_name)
```

